# Practical Machine Learning- Course Project 
Khansa Irem
April, 2018

# Introduction 
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement—a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# Data Processing 
## Load the Data
In this section, load the data and the 20 cases that will be submitted to coursera.

```{r echo=TRUE, message=FALSE, warning=FALSE}

library(ggplot2)
library(caret)
library(klaR)
library(rpart)
library(gbm)
library(randomForest)

rm(list = ls())
if (!file.exists("pml-training.csv")) {
  download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "pml-training.csv")
}
if (!file.exists("pml-testing.csv")) {
  download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "pml-testing.csv")
}
submit <- read.csv("pml-testing.csv", sep = ",", na.strings = c("", "NA"))
data <- read.csv("pml-training.csv", sep = ",", na.strings = c("", "NA"))
```

## Cleanup the data
Here, I remove columns full of NAs and remove features that are not in the submit set. The features containing NAs are the variance, mean and stddev within each window for each feature. Since the submit dataset has no time-dependence, these values are useless and can be disregarded. I also remove the first 7 features since they are related to the time-series or are not numeric.

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Remove columns full of NAs.
features <- names(submit[,colSums(is.na(submit)) == 0])[8:59]
# Only use features used in submit cases.
data <- data[,c(features,"classe")]
submit <- submit[,c(features,"problem_id")]
```

# Bootstrap
Next, I withhold 25% of the dataset for testing after the final model is constructed.

```{r echo=TRUE, message=FALSE, warning=FALSE}
set.seed(916)
inTrain = createDataPartition(data$classe, p = 0.75, list = F)
training = data[inTrain,]
testing = data[-inTrain,]
```

# Feature Selection
Some features may be highly correlated. The PCA method mixes the final features into components that are difficult to interpret; instead, I drop features with high correlation (>90%).

```{r echo=TRUE, message=FALSE, warning=FALSE}
outcome = which(names(training) == "classe")
highCorrCols = findCorrelation(abs(cor(training[,-outcome])),0.90)
highCorrFeatures = names(training)[highCorrCols]
training = training[,-highCorrCols]
outcome = which(names(training) == "classe")
```

# Feature Importance
The random forest method reduces overfitting and is good for nonlinear features. First, to see if the data is nonlinear, I use the random forest to discover the most important features. The feature plot for the 4 most important features is shown.

```{r echo=TRUE, message=FALSE, warning=FALSE}
fsRF = randomForest(training[,-outcome], training[,outcome], importance = T)
rfImp = data.frame(fsRF$importance)
impFeatures = order(-rfImp$MeanDecreaseGini)
inImp = createDataPartition(data$classe, p = 0.05, list = F)
featurePlot(training[inImp,impFeatures[1:4]],training$classe[inImp], plot = "pairs")
```

The most important features are:

**1.** pitch_belt
**2.** yaw_belt
**3.** total_accel_belt
**4.** gyros_belt_x

# Training
Train using the random forest and k-nearest neighbors for comparison.

```{r echo=TRUE, message=FALSE, warning=FALSE}
ctrlKNN = trainControl(method = "adaptive_cv")
modelKNN = train(classe ~ ., training, method = "knn", trControl = ctrlKNN)
ctrlRF = trainControl(method = "oob")
modelRF = train(classe ~ ., training, method = "rf", ntree = 200, trControl = ctrlRF)
resultsKNN = data.frame(modelKNN$results)
resultsRF = data.frame(modelRF$results)
```

# Testing Out-of-sample error
The random forest will give a larger accuracy compared to k-nearest neighbors. Here, I give the confusion matrix between the KNN and RF models to see how much they agree on the test set, then I compare each model using the test set outcomes.

```{r echo=TRUE, message=FALSE, warning=FALSE}
fitKNN = predict(modelKNN, testing)
fitRF = predict(modelRF, testing)
```

## KNN vs. RF

 Confusion Matrix and Statistics
 
           Reference
 
 Prediction    A    B    C    D    E
 
          A 1360    8   12   16    7

          B   37  834   27   24   25

          C   14   31  780   22    9

          D    9    3   56  721   14

          E   14   33   22   36  790
 
 Overall Statistics
                                         
                Accuracy : 0.915         
                  95% CI : (0.906, 0.922)
     No Information Rate : 0.292         
     P-Value [Acc > NIR] : < 2e-16       
                                         
                   Kappa : 0.892         
  Mcnemar's Test P-Value : 3.1e-11       
 
 Statistics by Class:
 -------------------  -------- -------  -------  -------- --------
                      Class: A Class: B Class: C Class: D Class: E
                      
 Sensitivity             0.948    0.917    0.870    0.880    0.935
 
 Specificity             0.988    0.972    0.981    0.980    0.974
 
 Pos Pred Value          0.969    0.881    0.911    0.898    0.883
 
 Neg Pred Value          0.979    0.981    0.971    0.976    0.986
 
 Prevalence              0.292    0.185    0.183    0.167    0.172
 
 Detection Rate          0.277    0.170    0.159    0.147    0.161
 
 Detection Prevalence    0.286    0.193    0.175    0.164    0.183
 
 Balanced Accuracy       0.968    0.945    0.925    0.930    0.955
 
 -------------------    ------   ------   ------   ------- -------

## KNN vs. test set

 Confusion Matrix and Statistics
 
           Reference
 Prediction    A    B    C    D    E
 
          A 1354   44   13    9   14
          
          B    6  835   32    3   33
          
          C   12   25  779   59   22
          
          D   16   24   20  723   36
          
          E    7   21   11   10  796
 
 Overall Statistics
                                         
                Accuracy : 0.915         
                  95% CI : (0.907, 0.923)
     No Information Rate : 0.284         
     P-Value [Acc > NIR] : < 2e-16       
                                         
                   Kappa : 0.892         
  Mcnemar's Test P-Value : 3.91e-15      
 
 Statistics by Class:
 -------------------  -------- -------  -------  -------- --------
                      Class: A Class: B Class: C Class: D Class: E
                      
 Sensitivity             0.971    0.880    0.911    0.899    0.883
 
 Specificity             0.977    0.981    0.971    0.977    0.988
 
 Pos Pred Value          0.944    0.919    0.868    0.883    0.942
 
 Neg Pred Value          0.988    0.971    0.981    0.980    0.974
 
 Prevalence              0.284    0.194    0.174    0.164    0.184
 
 Detection Rate          0.276    0.170    0.159    0.147    0.162
 
 Detection Prevalence    0.292    0.185    0.183    0.167    0.172
 
 Balanced Accuracy       0.974    0.931    0.941    0.938    0.936
-------------------    ------   -------  -------   ------- -------

## RF vs. test set

  Confusion Matrix and Statistics
 
           Reference
 Prediction    A    B    C    D    E
 
          A 1395    8    0    0    0
          
          B    0  940    7    0    0
          
          C    0    1  845   10    0
          
          D    0    0    3  794    6
          
          E    0    0    0    0  895
 
 Overall Statistics
                                        
                Accuracy : 0.993        
                  95% CI : (0.99, 0.995)
     No Information Rate : 0.284        
     P-Value [Acc > NIR] : <2e-16       
                                        
                   Kappa : 0.991        
  Mcnemar's Test P-Value : NA           
 
 Statistics by Class:
 -------------------  -------- -------  -------  -------- --------
                      Class: A Class: B Class: C Class: D Class: E
                      
 Sensitivity             1.000    0.991    0.988    0.988    0.993
 
 Specificity             0.998    0.998    0.997    0.998    1.000
 
 Pos Pred Value          0.994    0.993    0.987    0.989    1.000
 
 Neg Pred Value          1.000    0.998    0.998    0.998    0.999
 
 Prevalence              0.284    0.194    0.174    0.164    0.184
 
 Detection Rate          0.284    0.192    0.172    0.162    0.183
 
 Detection Prevalence    0.286    0.193    0.175    0.164    0.183
 
 Balanced Accuracy       0.999    0.994    0.993    0.993    0.997
 -------------------    ------   -------  -------   ------- -------

The random forest fit is clearly more accurate than the k-nearest neighbors method with 99% accuracy.



